The not so widening achievement gap: an interntational comparison of the evolution of the SES achievement gap from 2000-2015

### 2017-08-28 paper/paper.Rnw
included bayesian modelling - new specification of models - final model

<<default_conf>>=
library(knitr)
library(arm)
library(saves)
library(haven)
library(PISA2000lite)
library(PISA2003lite)
library(PISA2006lite)
library(PISA2009lite)
library(PISA2012lite)
library(intsvy)
library(cimentadaj) # devtools::install_github("cimentadaj/cimentadaj")
library(countrycode) # For region variable
library(car)
library(SAScii)
library(inequalityintsvy) # devtools::install_github("cimentadaj/inequalityintsvy")
library(lme4)
library(modelr)
library(ggrepel)
library(xtable)
library(ggthemes)
library(matrixStats)
library(WDI)
library(rstanarm)
library(tidyverse)

@@

  pisa_all2$value[[6]] <-
    pisa_all2$value[[6]] %>%
    rename(escs_trend = escs_trend)
@

@@

# results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.25, 0.75))
# results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.25, 0.75))
# results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.75))
# results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.75))
# results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.25, 0.5))
# results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.25, 0.5))

# results_math_75 <- results_math
# results_read_75 <- results_read
# results_math_topmid_75 <- results_math_topmid
# results_read_topmid_75 <- results_read_topmid
# results_math_midbottom_75 <- results_math_midbottom
# results_read_midbottom_75 <- results_read_midbottom

results_math <- read_rds("./paper/data/delete.Rdata")
results_read <- read_rds("./paper/data/delete_read.Rdata")
results_math_topmid <- read_rds("./paper/data/delete_math_topmid.Rdata")
results_read_topmid <- read_rds("./paper/data/delete_read_topmid.Rdata")
results_math_midbottom <- read_rds("./paper/data/delete_math_midbottom.Rdata")
results_read_midbottom <- read_rds("./paper/data/delete_read_midbottom.Rdata")
# US is missing for reading

@@

escs_data <- escs_dummy_creator(adapted_year_data, c(0.25, 0.75))

@@

complete_data_topbottom <- mutate(complete_data_topbottom, type = "75th/25th SES gap")
complete_data_topmid <- mutate(complete_data_topmid, type = "75th/50th SES gap")
complete_data_midbottom <- mutate(complete_data_midbottom, type = "50th/25th SES gap")
@

@@

\begin{figure}
\begin{center}
<<graphing_9010gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
# 90/10 gaps across countries

df <- complete_data_topbottom

diff_increase_fun <- function(df) {

@@

p <- 
  complete_data_topbottom %>%
  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
  filter(type_test == "math", !is.na(country)) %>%
  ggplot(aes(as.character(wave), difference)) +
  geom_point() +
  facet_wrap(~ country, ncol = 5) +
  geom_linerange(aes(ymin = 0, ymax = difference))

p + geom_line(data = complete_data_topbottom %>%
                mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
                filter(!is.na(country)),
              stat = "smooth", method = "lm", aes(group = 1),
              formula = y ~ splines::ns(x, 2), size = 0.7,
              colour = "red") +
  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  ggthemes::theme_few() +
  theme(panel.spacing = unit(1, "lines"),
        panel.grid.major.y = element_line(colour = "grey")) +
  ggtitle("Evolution of the 90/10 achievement gap")

@
\caption{Evolution of the achievement gap for selected countries}
\end{center}
\end{figure}

<<>>=
m <- complete_data_topbottom %>%
  filter(type_test == "math") %>%
  select(wave, country, difference) %>%
  left_join(tracking_data, by = c("country" = "cntry_name")) %>%
  left_join(central_examination) %>%
  mutate(num_tracks = factor(tracks15y))

m %>%
  filter(country %in% countries) %>%
  split(.$country) %>%
  map(~ lm(difference ~ wave + ztrack, data = .x))

p + geom_smooth(data = complete_data_topbottom %>%
                mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
                filter(!is.na(country)),
                method = "lm", se = FALSE,
                aes(group = 1), size = 0.7,
                colour = "red") +
  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  ggthemes::theme_few() +
  theme(panel.spacing = unit(1, "lines"),
        panel.grid.major.y = element_line(colour = "grey")) +
  ggtitle("Evolution of the 90/10 achievement gap")
@

@@

@@

# 2015
# SC013Q01TA
# 1 Public
# 2 Private
# 7 N/A
# 8 Invalid
# 9 Missing

@@

difference_data <-
  complete_data_topbottom %>%
  filter(type_test == "math") %>%
  select(wave, country, difference) %>%
  rename(year = wave) %>%
  mutate(year = as.character(year))

@@

ready_data_age <- mutate(ready_data, age_selection = ifelse(selage >= 15, 1, 0))

library(brms)

mod_tracking <-
  brms::brm(
  difference ~
    num_tracks +
    age_selection + (1 | country),
  family = gaussian(),
  data = ready_data_age,
  prior = c(set_prior("cauchy(2, 0.5)", coef = "num_tracks1"),
            set_prior("cauchy(13, 1)", coef = "age_selection")
            ),
  warmup = 1000, iter = 2000, chains = 5
  )

@@

@@

pp_check(mod_tracking, type = "error_binned", nsamples = 1000)
# The models tends to overpredict high values and underpredict low values.
@

@@

vars_unique <- c("length", "num_tracks", "age_selection", "central_examination", "ztrack")

@@

mod_tracking <-
  ready_data_two %>%
  lm(cum_diff ~ age_selection, data = .) %>%
  arm::display(detail = T)

mod_tracking <-
  ready_data_two %>%
  filter(!is.na(age_selection)) %>%
  brm(
    cum_diff ~ age_selection,
    family = gaussian(),
    data = .,
    prior = set_prior("cauchy(13, 1)", coef = "age_selection"),
    warmup = 1000, iter = 2000, chains = 5
  )
pp_check(mod_tracking, type = "dens_overlay", nsamples = 100)
pp_check(mod_tracking, type = "ecdf_overlay", nsamples = 10)
pp_check(mod_tracking, type = "boxplot", nsamples = 30)
pp_check(mod_tracking, type = "stat", nsamples = 200) # very nice to compare central tendency

@@

>>>>

<<default_conf>>=
  library(matrixStats)
  library(tidyverse)
  library(knitr)
  library(arm)
  library(saves)
  library(haven)
  library(PISA2000lite)
  library(PISA2003lite)
  library(PISA2006lite)
  library(PISA2009lite)
  library(PISA2012lite)
  library(intsvy)
  library(cimentadaj) # devtools::install_github("cimentadaj/cimentadaj")
  library(countrycode) # For region variable
  library(car)
  library(SAScii)
  library(inequalityintsvy) # devtools::install_github("cimentadaj/inequalityintsvy")
  library(lme4)
  library(modelr)
  library(ggrepel)
  library(xtable)
  library(ggthemes)

@@

  pisa_all2$value[[6]] <-
    pisa_all2$value[[6]] %>%
    rename(escs_trend = ESCS)
@

@@

results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))

# results_math <- read_rds("./data/delete.Rdata")
# results_read <- read_rds("./data/delete_read.Rdata")
# results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
# results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
# results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
# results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
# US is missing for reading

@@

escs_data <- escs_dummy_creator(adapted_year_data, c(0.1, 0.9))

@@

complete_data_topbottom <- mutate(complete_data_topbottom, type = "90th/10th SES gap")
complete_data_topmid <- mutate(complete_data_topmid, type = "90th/50th SES gap")
complete_data_midbottom <- mutate(complete_data_midbottom, type = "50th/10th SES gap")
@

@@

\begin{figure}
\begin{center}
<<graphing_9010gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
# 90/10 gaps acros countries
diff_increase_fun <- function(df) {

@@

pooled_trendline <-
  complete_data_topbottom %>%
  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
  filter(!is.na(country))

pooled_trendline %>%
  filter(type_test == "math") %>%
  ggplot(aes(as.character(wave), difference)) +
  geom_point() +
  geom_linerange(aes(ymin = 0, ymax = difference))+
  geom_line(data = pooled_trendline, stat = "smooth", method = "lm", aes(group = 1),
            formula = y ~ splines::ns(x, 2), size = 0.7,
            colour = "red") +
  facet_wrap(~ country, ncol = 5) +
  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  ggthemes::theme_few() +
  theme(panel.spacing = unit(1, "lines"),
        panel.grid.major.y = element_line(colour = "grey")) +
  ggtitle("Evolution of the 90/10 achievement gap")
@
\caption{Evolution of the achievement gap for selected countries}
\end{center}
\end{figure}

@@

\pagebreak

\section{Appendix}

<<estimating_othergaps>>=
results_math_80 <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.8, 0.2))
results_read_80 <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.8, 0.2))
results_math_topmid_80 <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.8, 0.5))
results_read_topmid_80 <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.8, 0.5))
results_math_midbottom_80 <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.2))
results_read_midbottom_80 <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.2))

results_math_70 <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.70, 0.30))
results_read_70 <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.70, 0.30))
results_math_topmid_70 <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.70, 0.5))
results_read_topmid_70 <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.70, 0.5))
results_math_midbottom_70 <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.30))
results_read_midbottom_70 <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.30))

write_rds(results_math_80, "./data/math_80")
write_rds(results_read_80, "./data/read_80")
write_rds(results_math_topmid_80, "./data/math_topmid_80")
write_rds(results_read_topmid_80, "./data/read_topmid_80")
write_rds(results_math_midbottom_80, "./data/math_midbottom_80")
write_rds(results_read_midbottom_80, "./data/read_midbottom_80")

write_rds(results_math_70, "./data/math_70")
write_rds(results_read_70, "./data/read_70")
write_rds(results_math_topmid_70, "./data/math_topmid_70")
write_rds(results_read_topmid_70, "./data/read_topmid_70")
write_rds(results_math_midbottom_70, "./data/math_midbottom_70")
write_rds(results_read_midbottom_70, "./data/read_midbottom_70")

# results_math <- read_rds("./data/delete.Rdata")
# results_read <- read_rds("./data/delete_read.Rdata")
# results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
# results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
# results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
# results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")

# results_math <- read_rds("./data/delete.Rdata")
# results_read <- read_rds("./data/delete_read.Rdata")
# results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
# results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
# results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
# results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")

<<eval = F>>=
results_math_80 <- read_rds("./paper/data/math_80.Rdata")
results_read_80 <- read_rds("./paper/data/read_80.Rdata")
results_math_topmid_80 <- read_rds("./paper/data/math_topmid_80.Rdata")
results_read_topmid_80 <- read_rds("./paper/data/read_topmid_80.Rdata")
results_math_midbottom_80 <- read_rds("./paper/data/math_midbottom_80.Rdata")
results_read_midbottom_80 <- read_rds("./paper/data/read_midbottom_80.Rdata")

results_math_70 <- read_rds("./paper/data/math_70.Rdata")
results_read_70 <- read_rds("./paper/data/read_70.Rdata")
results_math_topmid_70 <- read_rds("./paper/data/math_topmid_70.Rdata")
results_read_topmid_70 <- read_rds("./paper/data/read_topmid_70.Rdata")
results_math_midbottom_70 <- read_rds("./paper/data/math_midbottom_70.Rdata")
results_read_midbottom_70 <- read_rds("./paper/data/read_midbottom_70.Rdata")
@

\begin{figure}
\begin{center}
<<graph_all_gaps,  include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
complete_data_topbottom_80 <-
  pisa_preparer(results_math_80, results_read_80) %>%
  mutate(type = "80th/20th SES gap")

complete_data_topmid_80 <-
  pisa_preparer(results_math_topmid_80, results_read_topmid_80) %>%
  mutate(type = "80th/50th SES gap")

complete_data_midbottom_80 <-
  pisa_preparer(results_math_midbottom_80, results_read_midbottom_80) %>%
  mutate(type = "50th/20th SES gap")

######

complete_data_topbottom_70 <-
  pisa_preparer(results_math_70, results_read_70) %>%
  mutate(type = "70th/30th SES gap")

complete_data_topmid_70 <-
  pisa_preparer(results_math_topmid_70, results_read_topmid_70) %>%
  mutate(type = "70th/50th SES gap")

complete_data_midbottom_70 <-
  pisa_preparer(results_math_midbottom_70, results_read_midbottom_70) %>%
  mutate(type = "50th/30th SES gap")

#####

complete_gaps <-
  complete_data_topbottom %>%
  bind_rows(complete_data_topbottom_80) %>%
  bind_rows(complete_data_topbottom_70)

complete_gaps %>%
  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
  filter(type_test == "math",
         country %in% countries) %>%
  ggplot(aes(as.character(wave), difference, colour = type, linetype = type, group = type)) +
  geom_line(stat = "smooth", method = "lm", size = 0.7) +
  facet_wrap(~ country, ncol = 5) +
  scale_colour_discrete(name = "Type of gaps") +
  scale_linetype_discrete(name = "Type of gaps") +
  scale_y_continuous(name = "Gaps in SD", expand = c(0, 0), lim = c(0, 3)) +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  ggthemes::theme_few() +
  theme(panel.spacing = unit(1, "lines"),
        panel.grid.major.y = element_line(colour = "grey"),
        legend.position = "bottom") +
  ggtitle("Evolution of the different achievement gaps")
@
\caption{Evolution of the achievement gap for several gaps}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
<<percentage_change_8020,  include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
perc_increase_fun(complete_data_topbottom_80) %>%
  perc_graph("math",
             "80/20 achievement gap",
             "Percentage change from 2000 to 2015")
@
\caption{Percentage change in the 80/20 achievement gap from 2000 to 2015}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
<<percentage_change_7030,  include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
perc_increase_fun(complete_data_topbottom_70) %>%
  perc_graph("math",
             "70/30 achievement gap",
             "Percentage change from 2000 to 2015")
@
\caption{Percentage change in the 70/30 achievement gap from 2000 to 2015}
\end{center}
\end{figure}

@@

# 2015
# SC013Q01TA
# 1 Public
# 2 Private
# 7 N/A
# 8 Invalid
# 9 Missing
library(readxl)

@@

difference_data <-
  complete_gaps %>%
  filter(type_test == "math", type == "90th/10th SES gap") %>%
  select(wave, country, difference) %>%
  rename(year = wave) %>%
  mutate(year = as.character(year))

@@

ready_data_age <- mutate(ready_data, age_selection = ifelse(selage >= 15, 1, 0) %>% as.factor)

library(brms)
# This linear model explains the pooled 90/10 differences
mod_tracking <-
  brms::brm(
    difference ~
      ztrack +
      zvoc +
      ztrack:zvoc, family = gaussian(),
  data = ready_data_age,
  warmup = 1000, iter = 2000, chains = 5
  )

mod_tracking
bayes_R2(mod_tracking)
marginal_effects(mod_tracking)

pp_check(mod_tracking, type = "dens_overlay", nsamples = 100)
pp_check(mod_tracking, type = "scatter_avg", nsamples = 100)
pp_check(mod_tracking, type = "boxplot", nsamples = 20)
pp_check(mod_tracking, type = "error_hist", nsamples = 6)
pp_check(mod_tracking, type = "ecdf_overlay")

pp_check(mod_tracking, type = "error_scatter_avg_vs_x", x = "zvoc")
pp_check(mod_tracking, type = "error_scatter_avg_vs_x", x = "ztrack")

# Check whether the previous model can be generalized to a multilevel
# and see how much it can explain between country var
mod_tracking <-
  brms::brm(
    difference ~
      ztrack +
      zvoc +
      ztrack:zvoc + (1 | country),
  family = gaussian(),
  data = ready_data_age,
  prior = c(set_prior("normal(0, 1)", coef = "ztrack"),
            set_prior("normal(0, 1)", coef = "zvoc")),
  warmup = 1000, iter = 2000, chains = 4
  )

@@

# The code bloew needs to be adapted to the model from above (
# this is from previous models)

@@

pp_check(mod_tracking, type = "error_binned", nsamples = 1000)
@

@@

vars_unique <- c("length", "num_tracks", "age_selection", "central_examination", "ztrack",
                 "zvoc")

@@

mod_tracking <-
  ready_data_two %>%
  lm(cum_diff ~ ztrack + zvoc, data = .) %>%
  arm::display(detail = T)

mod_tracking <-
  ready_data_two %>%
  filter(!is.na(age_selection)) %>%
  brm(
    cum_diff ~ ztrack + zvoc,
    family = gaussian(),
    data = .,
    warmup = 1000, iter = 2000, chains = 5
  )

pp_check(mod_tracking, type = "dens_overlay", nsamples = 100)
pp_check(mod_tracking, type = "ecdf_overlay", nsamples = 10)
pp_check(mod_tracking, type = "boxplot", nsamples = 30)
pp_check(mod_tracking, type = "stat", nsamples = 200) # very nice to compare central tendency

@@

# update all of this for the previous model!
y_rep <- posterior_predict(mod_tracking)
